---
title: "Dissimilarity Analysis - Hogup Cave & Crescent Springs"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

```{r setup, include=FALSE}
############################################################
# WORKFLOW FOR HOGUP–WETLAND DISSIMILARITY ANALYSIS
# Using Martin et al. (2017) phases, ordered Monte Carlo ages,
# and two analyses for the Stratum 9 residual sample.
############################################################
if (!require("here")) install.packages("here")
library(here)
library(analogue)
```

## Occupation Phase Boundaries
```{r occphases}
## STEP 1.1: Define Martin et al. (2017) Phase boundaries ----
## These are the phase windows you will use for pollen ages.

phase_bounds <- list(
  I  = c(start = 9790, end = 6490),  # Strata 1–7
  II = c(start = 5840, end = 3330),  # Stratum 8
  III= c(start = 2870, end = 2760),  # Strata 9–10
  IV = c(start = 2610, end =  360)   # Strata 11–16
)
```

```{r maptooccphases}
## STEP 1.2: Map each cave pollen sample (row) to a Phase ----

hogup_meta <- read.csv(here("..", "data", "raw", "hogup_pollen.csv"), check.names = TRUE)

# Function to map stratum to phase label
stratum_to_phase <- function(stratum) {
  if (stratum >= 1  & stratum <= 7)  return("I")
  if (stratum == 8)                  return("II")
  if (stratum == 9 | stratum == 10)  return("III")
  if (stratum >= 11 & stratum <= 16) return("IV")
  return(NA_character_)
}

hogup_meta$phase <- vapply(hogup_meta$stratum, stratum_to_phase, character(1))
hogup_meta$phase <- factor(hogup_meta$phase, levels = c("I", "II", "III", "IV"))
```

```{r confirmmapping}
## STEP 1.3: Confirm mapping from phases to age windows ----

table(hogup_meta$stratum, hogup_meta$phase)
```

```{r timestep}
## STEP 1.4: Choose a global time step for interpolation ----

time_step <- 250   # time step in years; adjust if needed
```

## Ordered Monte Carlo Age Assignments
```{r ordersamples}
## STEP 2.1: Order samples within each phase by stratigraphy ----
## We assume lower stratum number = older.

# Separate pollen data from metadata
meta_cols   <- c("sample", "stratum", "phase")
cave_pollen <- hogup_meta[, !colnames(hogup_meta) %in% meta_cols, drop = FALSE]
hogup_meta  <- hogup_meta[,  meta_cols]

# Set row names to sample ID to enforce alignment
rownames(cave_pollen) <- hogup_meta$sample
rownames(hogup_meta)  <- hogup_meta$sample

# Sort by stratum within phase (lower stratum number = older)
hogup_meta  <- hogup_meta[order(hogup_meta$phase, hogup_meta$stratum), ]
cave_pollen <- cave_pollen[rownames(hogup_meta), ]

```

```{r countsamples}
## STEP 2.2: Count samples per phase ----

samples_per_phase <- table(hogup_meta$phase)
samples_per_phase
```

```{r generateages}
## STEP 2.3: Function to generate one set of ordered ages for all samples ----

generate_ordered_ages <- function(meta, phase_bounds) {
  # meta: hogup_meta with columns sample, stratum, phase
  # Returns: numeric vector of ages aligned with rows of 'meta'
  
  ages <- numeric(nrow(meta))   # placeholder
  
  # Loop over phases
  for (ph in names(phase_bounds)) {
    idx <- which(meta$phase == ph)           # indices in this phase
    if (length(idx) == 0) next               # skip if no samples
    n   <- length(idx)                       # number of samples in phase
    
    # Draw and sort proportions across the whole phase
    p   <- sort(runif(n, 0, 1))              # ordered U(0,1)

    # Map to ages using phase start and end
    start <- phase_bounds[[ph]]["start"]
    end   <- phase_bounds[[ph]]["end"]
    ages_phase <- start + p * (end - start)  # ordered ages within phase

    # Shuffle ages within each stratum so within-stratum order is random,
    # while between-stratum ordering (older stratum = older ages) is preserved
    strata_in_phase <- meta$stratum[idx]
    for (s in unique(strata_in_phase)) {
      s_pos <- which(strata_in_phase == s)
      ages_phase[s_pos] <- sample(ages_phase[s_pos], size = length(s_pos))
    }

    # Assign to these rows (meta is sorted by stratum within phase)
    ages[idx] <- ages_phase
  }
  
  return(ages)
}
```

```{r residualsamples}
## STEP 2.4: Handle the Stratum 9 residual pollen sample ----

hogup_meta$is_residual <- FALSE
hogup_meta$is_residual[hogup_meta$sample == 357] <- TRUE

residual_age_range <- c(4478, 4873)  # calibrated mid-Holocene range (cal BP)

# Variant 1: generate ages excluding the residual sample (Analysis A)
generate_ordered_ages_excl_resid <- function(meta, phase_bounds) {
  meta_no_resid <- meta[!meta$is_residual, ]
  ages_no_resid <- generate_ordered_ages(meta_no_resid, phase_bounds)
  ages <- rep(NA_real_, nrow(meta))
  ages[!meta$is_residual] <- ages_no_resid
  # residual sample intentionally has NA (excluded)
  return(ages)
}

# Variant 2: generate ages including residual as mid-Holocene (Analysis B)
generate_ordered_ages_incl_resid <- function(meta, phase_bounds, resid_range) {
  meta_no_resid <- meta[!meta$is_residual, ]
  ages_no_resid <- generate_ordered_ages(meta_no_resid, phase_bounds)
  ages <- rep(NA_real_, nrow(meta))
  ages[!meta$is_residual] <- ages_no_resid
  
  # assign residual sample a mid-Holocene age (either fixed or uniform)
  resid_idx <- which(meta$is_residual)
  if (length(resid_idx) == 1) {
    ages[resid_idx] <- runif(1, resid_range[1], resid_range[2])
  }
  
  return(ages)
}
```

```{r montecarlo}
## STEP 2.5: Generate many Monte Carlo realizations of ages ----

n_real <- 200   # number of Monte Carlo realizations (adjust as needed)

# Storage lists
age_realizations_excl <- vector("list", n_real)
age_realizations_incl <- vector("list", n_real)

set.seed(123)   # reproducibility

for (k in 1:n_real) {
  age_realizations_excl[[k]] <- generate_ordered_ages_excl_resid(hogup_meta, phase_bounds)
  age_realizations_incl[[k]] <- generate_ordered_ages_incl_resid(hogup_meta, phase_bounds,
                                                                 resid_range = residual_age_range)
}
```

Age vectors now align with rows of `cave_pollen`. For realization `k`, cave ages excluding the residual are in `age_realizations_excl[[k]]` and cave ages including the residual as mid-Holocene are in `age_realizations_incl[[k]]`.

## Dissimilarity Analysis with Crescent Springs Core
```{r loadwetdata}
## STEP 3.0: Load Crescent Springs wetland pollen data ----

crescent_raw <- read.csv(here("..", "data", "raw", "crescent_pollen.csv"), check.names = TRUE)

wet_age <- crescent_raw$Age  # cal yr BP

meta_cols  <- c("Depth", "Age", "Pollen.Sum", "Terrestrial", "Lyco")
wet_pollen <- crescent_raw[, !colnames(crescent_raw) %in% meta_cols, drop = FALSE]
rownames(wet_pollen) <- crescent_raw$Depth
```

```{r prepcsdata}
## STEP 3.1: Prepare wetland core and harmonize taxa ----

# Convert to proportions
wet_prop  <- as.data.frame(wet_pollen / 100)
cave_prop <- as.data.frame(cave_pollen / 100)

# Harmonize taxa
common_taxa <- intersect(colnames(wet_prop), colnames(cave_prop))
wet_prop    <- wet_prop[,  common_taxa]
cave_prop   <- cave_prop[, common_taxa]
```

```{r definetimegrid}
## STEP 3.2: Define a common time grid ----

# Use wetland age range and full Hogup phase ranges for bounding
age_min <- max(min(wet_age), min(unlist(lapply(phase_bounds, `[[`, "end"))))
age_max <- min(max(wet_age), max(unlist(lapply(phase_bounds, `[[`, "start"))))

time_grid <- seq(age_min, age_max, by = time_step)
```

```{r intpollenmatrix}
## STEP 3.3: Helper to interpolate a pollen matrix to time_grid ----

interp_pollen <- function(age_vec, prop_mat, new_ages) {
  # age_vec: vector of sample ages (cal BP)
  # prop_mat: rows = samples, cols = taxa (proportions)
  # new_ages: desired ages (time_grid)
  
  ord <- order(age_vec)
  age_vec  <- age_vec[ord]
  prop_mat <- prop_mat[ord, , drop = FALSE]
  
  interp_mat <- sapply(
    1:ncol(prop_mat),
    function(j) approx(x = age_vec,
                       y = prop_mat[, j],
                       xout = new_ages,
                       rule = 2)$y
  )
  interp_df <- as.data.frame(interp_mat)
  colnames(interp_df) <- colnames(prop_mat)
  rownames(interp_df) <- as.character(new_ages)
  
  # renormalize rows to sum to 1 (guard against all-zero rows)
  rs <- rowSums(interp_df)
  rs[rs == 0] <- 1
  interp_df <- sweep(interp_df, 1, rs, "/")
  
  return(interp_df)
}
```

## Squared Chord Distance & Rate of Change
```{r dissfunctions}
## STEP 3.4: Functions for squared chord distance and RoC ----

compute_roc <- function(prop_df, ages) {
  # prop_df rows must correspond to 'ages' (time_grid)
  scd <- analogue::distance(
    x = prop_df[-nrow(prop_df), , drop = FALSE],
    y = prop_df[-1,           , drop = FALSE],
    method = "SQchord"
  )
  age_mid <- (ages[-nrow(prop_df)] + ages[-1]) / 2
  data.frame(age_mid = age_mid, SCD = diag(scd))
}

compute_between <- function(wet_df, cave_df, ages) {
  # wet_df, cave_df on same time_grid
  scd <- analogue::distance(
    x = wet_df,
    y = cave_df,
    method = "SQchord"
  )
  data.frame(age = ages, SCD = diag(scd))
}

# Wetland interpolation and RoC are identical across all realizations
wet_interp <- interp_pollen(wet_age, wet_prop, time_grid)
roc_wet    <- compute_roc(wet_interp, time_grid)

# Run Monte Carlo dissimilarity analyses
# Storage: one list per realization
roc_cave_A <- vector("list", n_real)
between_A  <- vector("list", n_real)

roc_cave_B <- vector("list", n_real)
between_B  <- vector("list", n_real)

for (k in 1:n_real) {
  # Analysis A: exclude residual
  cave_age_A    <- age_realizations_excl[[k]]
  cave_interp_A <- interp_pollen(cave_age_A[!hogup_meta$is_residual],
                                 cave_prop[!hogup_meta$is_residual, ],
                                 time_grid)

  roc_cave_A[[k]] <- compute_roc(cave_interp_A, time_grid)
  between_A[[k]]  <- compute_between(wet_interp, cave_interp_A, time_grid)

  # Analysis B: include residual as mid-Holocene
  cave_age_B    <- age_realizations_incl[[k]]
  cave_interp_B <- interp_pollen(cave_age_B, cave_prop, time_grid)

  roc_cave_B[[k]] <- compute_roc(cave_interp_B, time_grid)
  between_B[[k]]  <- compute_between(wet_interp, cave_interp_B, time_grid)
}
```

## Summarize Monte Carlo Results
```{r summarize}
## STEP 3.5: Summarize across realizations ----

# Helper: stack one SCD column from all realizations into a matrix
stack_metric <- function(list_of_df, colname = "SCD") {
  do.call(cbind, lapply(list_of_df, `[[`, colname))
}

# Helper: compute median and 95% CI envelope from a stacked matrix
summarise_metric <- function(mat, ages, age_col = "age_mid") {
  data.frame(
    setNames(list(ages), age_col),
    median = apply(mat, 1, median,   na.rm = TRUE),
    q025   = apply(mat, 1, quantile, probs = 0.025, na.rm = TRUE),
    q975   = apply(mat, 1, quantile, probs = 0.975, na.rm = TRUE)
  )
}

age_mid_vec <- roc_cave_A[[1]]$age_mid  # shared midpoint ages (RoC metrics)
age_vec     <- between_A[[1]]$age       # shared time-grid ages (between metrics)

# Wetland RoC is deterministic (no age uncertainty); reshape to match other summaries
roc_wet_summary <- data.frame(
  age_mid = roc_wet$age_mid,
  median  = roc_wet$SCD,
  q025    = roc_wet$SCD,
  q975    = roc_wet$SCD
)

# Analysis A
roc_cave_A_summary <- summarise_metric(stack_metric(roc_cave_A), age_mid_vec, "age_mid")
between_A_summary  <- summarise_metric(stack_metric(between_A),  age_vec,     "age")

# Analysis B
roc_cave_B_summary <- summarise_metric(stack_metric(roc_cave_B), age_mid_vec, "age_mid")
between_B_summary  <- summarise_metric(stack_metric(between_B),  age_vec,     "age")
```

## Results

```{r fig1, fig.width=9, fig.height=5}
library(ggplot2)

fig1_dat <- rbind(
  cbind(between_A_summary, analysis = "A (excl. residual)"),
  cbind(between_B_summary, analysis = "B (incl. residual)")
)
fig1_dat$analysis <- factor(fig1_dat$analysis,
                             levels = c("A (excl. residual)", "B (incl. residual)"))

fig1 <- ggplot(fig1_dat, aes(x = age, y = median,
                              colour = analysis, fill = analysis)) +
  geom_ribbon(aes(ymin = q025, ymax = q975), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  scale_x_reverse(name = "Age (cal BP)",
                  breaks = seq(0, 10000, by = 1000)) +
  scale_colour_manual(values = c("A (excl. residual)" = "#E69F00",
                                 "B (incl. residual)" = "#56B4E9"),
                      name = "Analysis") +
  scale_fill_manual(values = c("A (excl. residual)" = "#E69F00",
                                "B (incl. residual)" = "#56B4E9"),
                    name = "Analysis") +
  labs(y = "Squared Chord Distance",
       title = "Between-site dissimilarity: Hogup Cave vs. Crescent Springs",
       caption = "Shaded band = 95% Monte Carlo envelope (n = 200 realizations)") +
  theme_bw(base_size = 11)

print(fig1)
```

**Figure 1** shows how dissimilar the pollen assemblages at Hogup Cave were from the contemporaneous wetland at Crescent Springs through time. High SCD indicates the two sites were recording very different vegetation — the cave and wetland assemblages were diverging. Low SCD indicates convergence, either because the wetland was influencing cave deposits or both sites were responding to the same regional conditions. Sustained high dissimilarity across a phase suggests Hogup Cave was accumulating pollen from a distinctly different local or regional source than the wetland (e.g., upland steppe taxa vs. wetland aquatics and halophytes). A drop in dissimilarity would suggest a convergence in vegetation — perhaps a wetter period pulling the cave assemblage toward wetland-type taxa — or a shift in human activity affecting what pollen entered the cave. The two analysis lines (A and B) allow assessment of whether the Stratum 9 residual sample meaningfully affects the pattern. If both lines track closely, the residual has little influence and the interpretation is robust. If they diverge, the timing and placement of that sample is driving the result and warrants closer attention.

```{r fig2, fig.width=9, fig.height=5}
fig2_cave <- rbind(
  cbind(roc_cave_A_summary, source = "Cave (A, excl. residual)"),
  cbind(roc_cave_B_summary, source = "Cave (B, incl. residual)")
)
fig2_cave$source <- factor(fig2_cave$source,
                            levels = c("Cave (A, excl. residual)",
                                       "Cave (B, incl. residual)"))

fig2_wet <- roc_wet_summary  # deterministic — single line, no envelope

fig2 <- ggplot() +
  geom_line(data = fig2_wet,
            aes(x = age_mid, y = median),
            colour = "#009E73", linetype = "dashed", linewidth = 0.8) +
  geom_ribbon(data = fig2_cave,
              aes(x = age_mid, ymin = q025, ymax = q975,
                  fill = source),
              alpha = 0.2) +
  geom_line(data = fig2_cave,
            aes(x = age_mid, y = median, colour = source),
            linewidth = 0.8) +
  scale_x_reverse(name = "Age (cal BP)",
                  breaks = seq(0, 10000, by = 1000)) +
  scale_colour_manual(values = c("Cave (A, excl. residual)" = "#E69F00",
                                 "Cave (B, incl. residual)" = "#56B4E9"),
                      name = "Cave analysis") +
  scale_fill_manual(values = c("Cave (A, excl. residual)" = "#E69F00",
                                "Cave (B, incl. residual)" = "#56B4E9"),
                    name = "Cave analysis") +
  annotate("text", x = max(roc_wet_summary$age_mid), y = max(roc_wet_summary$median),
           label = "Crescent Springs (wetland)", colour = "#009E73",
           hjust = 0, vjust = -0.3, size = 3) +
  labs(y = "Squared Chord Distance",
       title = "Within-site rate of change",
       caption = "Shaded band = 95% Monte Carlo envelope (n = 200 realizations).\nWetland RoC is deterministic (no age uncertainty).") +
  theme_bw(base_size = 11)

print(fig2)
```

**Figure 2** shows how rapidly the pollen assemblage was changing internally at each site between consecutive time steps. High RoC (high SCD) indicates a period of ecological instability or rapid vegetation turnover, while low RoC indicates a stable, persistent vegetation community. The wetland line (dashed green) is deterministic — it carries no age uncertainty — so its peaks represent real, dateable vegetation transitions in the Crescent Springs basin. The cave envelopes (orange/blue) carry Monte Carlo age uncertainty: wider bands indicate time periods with fewer or more loosely constrained samples, where the biological signal is real but the exact timing is uncertain. Comparing the two sites, if a RoC peak appears at roughly the same time in both the cave and the wetland, it suggests a synchronous regional event such as a climatic shift. Asynchronous peaks may reflect site-specific processes — changes in human occupation intensity, local microenvironment, or taphonomy at the cave. As with Figure 1, comparing lines A and B reveals whether the Stratum 9 residual affects the timing or magnitude of rate-of-change events.

### Summary tables

```{r tables}
knitr::kable(between_A_summary, digits = 4,
             col.names = c("Age (cal BP)", "Median SCD", "2.5%", "97.5%"),
             caption = "Between-site SCD — Analysis A (residual excluded)")

knitr::kable(between_B_summary, digits = 4,
             col.names = c("Age (cal BP)", "Median SCD", "2.5%", "97.5%"),
             caption = "Between-site SCD — Analysis B (residual as mid-Holocene)")

knitr::kable(roc_cave_A_summary, digits = 4,
             col.names = c("Age midpoint (cal BP)", "Median SCD", "2.5%", "97.5%"),
             caption = "Cave RoC — Analysis A")

knitr::kable(roc_cave_B_summary, digits = 4,
             col.names = c("Age midpoint (cal BP)", "Median SCD", "2.5%", "97.5%"),
             caption = "Cave RoC — Analysis B")

knitr::kable(roc_wet_summary, digits = 4,
             col.names = c("Age midpoint (cal BP)", "Median SCD", "2.5%", "97.5%"),
             caption = "Wetland RoC — Crescent Springs (deterministic)")
```

